Demystifying the Computer Science Ph.D. Admission in the US

### 2023-06-10 phd-cs-us.tex
Updates from Overleaf

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: One of the reasons I create this post is that several of my colleagues are interested in 
recruiting Vietnamese students and were surprised when seeing very few applications in Vietnam (e.g., our CS PhD program has on average 3 applications from Vietnam each year). In general the number of
PhD applications from Vietnam to US universities is very low and  more would be very welcomed. 
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: We usually decide that a candidate is either (i) admit with funding (TA or RA) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we don't. In some rare cases, we may admit
you with no funding because you have funding on your own (e.g., you are
supported by your government or have external grants). Also, we justify
our decision with a short summary about your application, where we list
strengths (e.g., well-known undergrad school) and weaknesses (e.g., weak
LORs).
\end{tcolorbox}

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
    Hakan: At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline (December 1st), and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as "optional", it is mandatory and they should plan for submitting all those.  
\end{tcolorbox}
\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments/programs do not have the flexibility to waive the application fee, even if they want to.

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: Vietnamese students often mention Scopus Q1, which consists of diverse journals from IEEE, Elsevier, and several other publishers unfamiliar to me.  Honestly, I don't know/recognize the majority of journals listed in Scopus Q1. So this might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed in there, so devote sometime in your statement to discuss the significance of your papers.
\end{tcolorbox}

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Craig: GMU and many other universities allow you to upload your published papers and other writing samples. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.  
\end{tcolorbox}

@@

\subsection{Research Statement}\label{sec:research-statement}

While you might not be able to get good LORs or change
\hyperref[sec:your-school]{where your go to school}, you have control of your
statement! So write it well because we do take it seriously.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

@@

Finally, this is something very easy to do, but is missed by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply here? talk about a couple of professors who you're interested in working with (in many cases your application will be forwarded to them for evaluation).
Be careful not to send wrong statement to wrong school or mixing
facts (e.g., talking about school X but mentioned about working with
profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
should.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: I always read the research statement first and then LORs. If I am
persuaded by then, I would skim over other factors and advocate for
admission (unless I see red flags in other parts). If I am not
convinced, then I will likely recommend rejection (unless I see
something standout in other parts).
\end{tcolorbox}

>>>>

\newtcolorbox{cbox}[1][]
{
left=1pt,right=1pt,top=1pt,bottom=1pt,
colback=gray!10,
boxrule=0.5pt,
title=#1 % I would like to make this (one of these in general) assignment optional depending on #1, #2...
}

 \newenvironment{commentbox}[1][]{
 \small
    \begin{cbox}[#1]
 }{
   \end{cbox}
}

@@

\begin{commentbox}[]
Vu: One of the reasons I create this post is that several of my colleagues are interested in 
recruiting Vietnamese students and were surprised when seeing very few applications in Vietnam (e.g., our CS PhD program has on average 3 applications from Vietnam each year). In general the number of
PhD applications from Vietnam to US universities is very low and  more would be very welcomed. 
\end{commentbox}

@@

\begin{commentbox}[]
Vu: We usually decide that a PhD candidate is either (i) admit with funding (TA or RA) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we don't. In some rare cases, we may admit
you with no funding because you have funding on your own (e.g., you are
supported by your government or have external grants). We justify
our decision with a short summary about your application, where we list
strengths (e.g., well-known undergrad school) and weaknesses (e.g., weak
LORs).
\end{commentbox}

\begin{commentbox}[]
    Hakan: At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as "optional", it is mandatory and they should plan for submitting all those.  
\end{commentbox}
\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments/programs do not have the flexibility to waive the application fee, even if they want to.

@@

\begin{commentbox}[]
Vu: Vietnamese students often mention Scopus Q1, which consists of diverse journals from IEEE, Elsevier, and many other publishers unfamiliar to me.  Honestly, I don't know/recognize many of journals listed in Scopus Q1 and. So this might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed in there, so devote sometime in your statement to discuss the significance of your papers.
\end{commentbox}

\begin{commentbox}[]
Craig: GMU and many other universities allow you to upload your published papers and other writing samples. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.  
\end{commentbox}

@@

\subsection{Research or Personal Statement}\label{sec:research-statement}

While you might not have control over LORs \hyperref[sec:your-school]{where your go to school}, you do over your
statement! So write it well because we do take it seriously.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

@@

Finally, this is something easy to do, but is missed by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply here? talk about a couple of professors who you're interested in working with (in many cases your application will be forwarded to them for evaluation).
This shows that you're serious and have done homework on where you're applying to.
Be careful not to send wrong statement to wrong school or mixing
facts (e.g., talking about school X but mentioned about working with
profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
should.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
\small
Vu: I always read the research statement first and then LORs. If I am
persuaded by then, I would skim over other factors and advocate for
admission (unless I see red flags in other parts). If I am not
convinced, then I will likely recommend rejection (unless I see
something standout in other parts).
\end{tcolorbox}
